### 2022-02-16 introduction/introduction.tex
more

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. For example, suppose we want to prove that there is a
function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
function can be represented using $\lambda$ abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stage is to support automatically synthesizing such functions.

@@

The biggest contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend those provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term structure of Ehoh to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and less redundant than current state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented the new calculus and the described tweaks. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.

  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining prover we call
  \newehoh{}. We again modified term structure, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, strong basis that we have earlier built made it
  manageable.

  \item Following our approach of lifting efficient provers from
  first-order to higher-order logic, we went a step back in an effort to improve
  first-order provers. In particular, we looked at the most efficient approaches
  to simplify problems in propositional logic, a simpler logic that disallows quantification,
  but it is decidable. We lifted some of these approaches to first-order logic
  and implemented them in Zipperposition. 
\end{itemize}

@@

The e
arliest attempts to automate higher-order logic can be traced back to
Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
logic to first-order logic using combinators, similar to the approach taken by
hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
calculi that are directly applied to higher-order formulas.

These attempts resulted in early higher-order theorem provers. Andrews and
collegaues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
and colleageues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced cooperative paradigm. Provers of this paradigm
invoke first-order back-ends in regular intervals to finish the proof. Latest
iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
\cite{sb-21-leo3} is based on higher-order analytic tableau and uses SAT solver
to close the branches. It won of higher-order division of CASC on 7 ocassions:
in 2011 and every year from 2013 to 2019 (and once in 2011).

Various reserachers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers of the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

Higher-order unification is one of the central procedures in a higher-order
prover. Jensen and Pietrzykowski introduced a complete procedure for enumerating
elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
some calculi remaing complete if solving a difficult subproblem of higher-order
unification (`flex-flex'' pairs) is delayed. He described procedure to solve
this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
desgined a procedure to enumerate higher-order unifiers using higher-order
combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat
and Reger implemented efficient term indexing in their extension of Vampire
\cite{br-20-full-sup-w-combs}. Libal and Steen designed efficient term indexing
based on substitution trees \cite{ls-16-indexing}.

>>>>

Higher-order logic also supports expressing functions that are not explicitly
present in the theory. Suppose we want to prove that there is a
function that returns 4 when given 2 and 5 when given 3 ($\exists x. x(\cst{2})
\eq \cst{4} \lland x (\cst{3}) \eq \cst{5} $). An example of such function is
the function that adds 2 to its argument. In higher-order logic such unnamed
function can be represented using $\lambda$-abstractions as $\lambda y.\, y \,
\cst{+} \, \cst{2}$. This term suffices to prove the original statement. The
second stage is to support automatically synthesizing such functions.

@@

The biggest contribution of this thesis is the extension of two theorem provers,
E and Zipperposition \cite{sc-15-simon-phd}, to full higher-order logic. To
extend those provers, we faced many challenges. Some of them were of engineering
nature, some of them were algorithmic, while others concerned fine-tuning the
heuristics to achieve best possible performance. We discuss those challenges in more detail:
\begin{itemize}
  \item We implemented the complete superposition calculus for $\lambda$-free
  higher-order logic \cite{bbcw-21-lfho} in E, obtaining a prover we call Ehoh.
  We had to modify internal term representation of Ehoh to support $\lambda$-free
  higher-order terms, implement new unification and matching algorithms, extend
  term indexing data structures to work with higher-order terms, and to improve the
  performance of heuristics on higher-order problems.
  \item Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second stage described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in short amount of time we decided to
  implement the calculus in Zipperposition. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition with Zipperposition
  \cite{gs-19-casc27}. It took third place, closely behind (1 percent
  point) Leo-III \cite{sb-21-leo3}  and far behind (12 percent points) Satallax \cite{cb-12-satallax}.
  \item Through this experience we identified one of the bottlenecks of
  Zipperposition: its unification algorithm. To remove this bottleneck we
  developed a complete higher-order unification algorithm that was more
  efficient and less redundant than current state of the art. 
  \item We also noticed that Zipperposition's support for first-class Booleans
  was lacking compared to competition. We have closely studied the problems on
  which Zipperposition fails, but are easily solved by other provers and created
  a set of incomplete rules that enhance Zipperposition's Boolean support.
  Zipperposition 2 implemented these rules together with the new unification
  algorithm. In 2020, Zipperposition 2 was the winner of higher-order division
  of CASC \cite{gs-21-cascj10}. This time it was 20 percent points better than the second best
  prover, Satallax.
  \item Inspired by this success we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third stage) that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}. 
  \item Taking all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has larger scope than the original
  extension from E to Ehoh, strong basis that we have earlier built made it
  manageable.

  \item Following our approach of lifting efficient provers from
  first-order to higher-order logic, we went a step back in an effort to improve
  first-order provers. In particular, we looked at the most efficient approaches
  to simplify problems in propositional logic, a simpler, decidable logic that disallows quantification. 
  We lifted some of these approaches to first-order logic
  and implemented them in Zipperposition. 
\end{itemize}

@@

The earliest attempts to automate higher-order logic can be traced back to
Robinson~\cite{ar-70-hol}. His approach was based on translating higher-order
logic to first-order logic using combinators, similar to the approach taken by
hammers. Huet~\cite{gh-73-hol} and Andrews~\cite{pa-71-type-theory} designed
calculi that are directly applied to higher-order formulas.

These attempts resulted in early higher-order theorem provers. Andrews and
collegaues developed TPS~\cite{abinpx-96-tps}: an automatic prover which
allows user to specify proof outlines, based on expansion proofs. Benzm\"uller
and colleageues developed LEO \cite{cbmk-98-leo}, a prover based on higher-order
resolution which introduced cooperative paradigm. Provers of this paradigm
invoke first-order back-ends in regular intervals to finish the proof. Latest
iteration in LEO family of provers is Leo-III \cite{sb-21-leo3}. Satallax
\cite{cb-12-satallax} is based on higher-order analytic tableau and uses SAT solver
to close tableau branches. It won higher-order division of CASC on 7 ocassions:
in 2011 and every year from 2013 to 2019.

Various reserachers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

Higher-order unification is one of the central procedures in a higher-order
prover. Jensen and Pietrzykowski introduced a complete procedure for enumerating
elements unifiers for higher-order terms \cite{jp-76-unif}. Huet noticed that
some calculi remaing complete if solving a difficult subproblem of higher-order
unification (``flex-flex'' pairs) is delayed. He described procedure to solve
this easier problem in a more efficient way \cite{gh-75-unification}. Dougherty
desgined a procedure to enumerate higher-order unifiers using higher-order
combinators~\cite{dd-93-comb-unif}. To make unification more efficient, Bhayat
and Reger implemented efficient term indexing in their extension of Vampire
\cite{br-20-full-sup-w-combs}. Libal and Steen designed efficient term indexing
based on substitution trees \cite{ls-16-indexing}.
